Income Has Serious Implications On Many Facets of American Life in 2021

### 2022-03-22 outputs/paper/paper.Rmd
tenth assist

# Data
In this Data Section \@ref(data), we will provide a look into the data acquisition and processing methodology as well as a deep dive into the contents of the data. First of, (Table \@ref(tab:tableofdata)) give us a glimpse of the data.
```{r echo = F}
#table of variables that i am concerned with
data_table <- reduced_data[,-c(1,6,8)]

data_table[complete.cases(data_table), ] %>%
   slice(1:7) %>%
  kable(
    caption = "GSS Cleaned Data",
    col.names = c("Age", "Gender", "Born", 
                  "Education", 
                  "Birth Year", "Grandparents", 
                  "Income at 16","Current Income"),
    digits = 1,
    booktabs = TRUE,  
    linesep = "")

@@

## Key features & Pros And Cons of the Methodology and Survey

It is important to discuss the reach of our data. We know from the provided documentation that the population group eligible for taking the survey are adults 18 or older in the United States who live in non institutional housing at the time of interviewing. Also, the 2021 GSS  uses the last birthday method to decide which individual in the household will take the survey. This was done by having "a professional phone interviewer team conducted phone outreach to complete screeners, answer inbound phone calls from sampled households, identify respondents, and complete interviews by phone and/or prompt them to complete the web survey." As far as the invitation method goes for data collection, materials that provide a web link to invite people to participate in the survey was mailed out and an option to conduct the interview via the phone instead of a self-administered questionnaire was also provided. Out of the 27,591 lines of sample (people who had been prompted to answer the questionnaire), 4,032 responses were collected. This has implications on the finding of our analysis in Section \@ref(limitations) as we will have to consider if we are able to generalize our finding from the relatively small sample that we have to the population of all US citizens over 18 years old. The small sample size and extremely long survey are both weaknesses of the survey which we will discuss in more depth in Section \@ref(limitations).

@@

income16_gender/income_gender
```
In figure \@ref(fig:figure-2), we see the overlapping of the two graphs, the first of which is the same as figure \@ref(fig:figure-1) and the bottom one is the same in essence but with the information for current family income instead of at age 16. The first and most striking part of this visual is that for the current income, the highest threshold is $25,000 or more. We see that the both of our graphs are heavily skewed to wards higher incomes.wesuspect that this is mainly a production of the bin widths decided on for the analysis.

@@

## Visuals 3
```{r echo=FALSE, fig.cap="write a figure caption here", fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
#family income at 16 grouped by where individual was born

@@

# Limitations
In this limitations section,wewant to focus on the data which is provided to us and have a  look into why it might not be easy to take the implications mentioned above at face value.

## Uneven Bin width for Income Groups.
The first major flaw I've found in the data is that the bins that both the income at age 16 and current income are recorded in are not all the same size. For example, in the case of the current income variable, the the first bin has a size of \$1,000, the second bin has a width of \$1,999, the third returns to about \$1,000 and as we go higher up the echelon, we have seemingly random alterations in bin widths from \$1,000 to \$2,000 to \$5,000. This can cause serious problems in statistics and especially for purposes of analysis.

@@

## Why cut off income at $25,000?
According to a news release published by the United States Bureau of Labor Statistics (see the document called wkyeng.pdf in the literature folder in inputs), the median weekly income of US citizens in the 4th quarter of 2021 (furthest from the start of the pandemic) is $1,010. This information is from that of 116.3 million full-time wage and salary workers. If we do some basic mathematics to approximate the annual earning, we get a figure of over \$52,000 dollars per year. As such, we are unable to understand why the bins for current income have been designed as such. This is a major flaw in the technique used by the surveyers to collect the data. In section \@ref(next-steps)

## Next steps

Weaknesses and next steps should also be included.

>>>>

# Data
In this Data Section \@ref(data), we will provide a look into the data acquisition and processing methodology as well as a deep dive into the contents of the data. First of, (Table \@ref(tab:tableofdata)) give us a glimpse of the data.
```{r tableofdata, echo = F}
#table of variables that i am concerned with
data_table <- reduced_data[,-c(1,6,8)]

data_table[complete.cases(data_table), ] %>%
   slice(1:7) %>%
  kable(
    caption = "GSS Cleaned Data",
    col.names = c("Age", "Gender", "Born", 
                  "Education", 
                  "Birth Year", "Grandparents", 
                  "Income at 16","Current Income"),
    digits = 1,
    booktabs = TRUE,  
    linesep = "")
```

@@

## Key features Pros And Cons of the Methodology and Survey

It is important to discuss the reach of our data. We know from the provided documentation that the population group eligible for taking the survey are adults 18 or older in the United States who live in non institutional housing at the time of interviewing. Also, the 2021 GSS  uses the last birthday method to decide which individual in the household will take the survey. This was done by having "a professional phone interviewer team conducted phone outreach to complete screeners, answer inbound phone calls from sampled households, identify respondents, and complete interviews by phone and/or prompt them to complete the web survey." As far as the invitation method goes for data collection, materials that provide a web link to invite people to participate in the survey was mailed out and an option to conduct the interview via the phone instead of a self-administered questionnaire was also provided. Out of the 27,591 lines of sample (people who had been prompted to answer the questionnaire), 4,032 responses were collected. This has implications on the finding of our analysis in Section \@ref(limitations) as we will have to consider if we are able to generalize our finding from the relatively small sample that we have to the population of all US citizens over 18 years old. The small sample size and extremely long survey are both weaknesses of the survey which we will discuss in more depth in Section \@ref(small-sample-size).

@@

income16_gender/income_gender
```
In figure \@ref(fig:figure2), we see the overlapping of the two graphs, the first of which is the same as figure \@ref(fig:figure-1) and the bottom one is the same in essence but with the information for current family income instead of at age 16. The first and most striking part of this visual is that for the current income, the highest threshold is $25,000 or more. We see that the both of our graphs are heavily skewed to wards higher incomes. We suspect that this is mainly a production of the bin widths decided on for the analysis.

@@

## Visuals 3
```{r, echo=FALSE, fig.cap="write a figure caption here", fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
#family income at 16 grouped by where individual was born

@@

# Limitations
In this limitations section,we want to focus on the data which is provided to us and have a  look into why it might not be easy to take the implications mentioned above at face value.

## Uneven Bin width for Income Groups.
The first major flaw I've found in the data is that the bins that both the of the current income variable, the the first bin has a size of \$1,000, the second bin has a width of \$1,999, the third returns to about \$1,000 and as we go higher up the echelon, we have seemingly random alterations in bin widths from \$1,000 to \$2,000 to \$5,000. This can cause serious problems in statistics and especially for purposes of analysis.

@@

## Why cut off income at $25,000?
According to a news release published by the United States Bureau of Labor Statistics (see the document called wkyeng.pdf in the literature folder in inputs), the median weekly income of US citizens in the 4th quarter of 2021 (furthest from the start of the pandemic) is $1,010. This information is from that of 116.3 million full-time wage and salary workers. If we do some basic mathematics to approximate the annual earning, we get a figure of over \$52,000 dollars per year. As such, we are unable to understand why the bins for current income have been designed as such. This is a major flaw in the technique used by the surveyors to collect the data. In section \@ref(next-steps), we will discuss ways that the survey could improve on this.

## Small sample size
As we have reported in Section \@ref(key-features-pros-and-cons-of-the-methodology-and-survey), out of the 27,591 lines of sample (people who had been prompted to answer the questionnaire), 4,032 responses were collected. This means that for the 2021 GSS, the response rate was at 17.4 percent (see calculation on page 28 of the "GSS 2021 Codebook R1b.pdf"). The United States has a population of over 330 million and even if all the people who had been prompted to answer the survey had responded it, the survey would only represent less than $8.36 \times 10^{-3}$ percent of the population. It is quite reasonable that as such, it would be hard to make generalizations for the entire adult united states population.

## Next steps
Following on from the latter section, we can make some pretty good suggestions on ways to improve the survey and subsequently, the data and the implications of the findings, so that they will be less ambiguous and more ethical to use with confidence. First off, having the data in predetermined bins is problematic from the perspective of the data analyst. We can start by creating even bins that respondents can answer but we can also allow the respondents to provide their exact income. The latter solution will give us more data, and it gives more options to the data analysis to conduct a reproducible analysis. In this manner, the data scientist who gets a hold of the data will have the liberty to conduct his analysis and chose the bins which work best for the task he is working on, instead of being obstructed by barriers built into the data.
